---
title: "Parameter Fitting with Synthetic Datasets - All Variants"
author: "Omid Arhami"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter Fitting with Synthetic Datasets - All Variants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  warning = TRUE,
 message = TRUE,    # Show messages
  error = TRUE      # Continue on error
)
```

## Introduction

This vignette demonstrates parameter fitting for synthetic datasets across different:
* Dimensionalities (2D, 5D, 10D)
* Missing data patterns (S, M, L levels)
* Noise types (random noise & biased noise)

There are 27 combinations. Therefore, to run the simulation in this script, local parallel processing will be used.

## Required Packages

```{r}
library(topolow)
library(ggplot2)
library(dplyr)
#library(data.table)
library(gridExtra)
```

## Generate Synthetic Datasets

First, let's create synthetic datasets with different dimensionalities:

```{r}
# Generate datasets in 2, 5, and 10 dimensions
results <- generate_synthetic_datasets(
  n_dims_list = c(2, 5, 10),
  seeds = c(123450, 903450, 13450),
  n_points = 250,
  output_dir = "synthetic_data"
)

# Print dataset properties
print(results$metadata)

# List available matrix variants for each dimension
matrix_variants <- names(results$matrices[[1]])[c(3,4,6,7,8,10,11,12,14)]
print("Available matrix variants:")
print(matrix_variants)
```

## Parameter Optimization Setup

Define consistent parameter ranges and optimization settings:

```{r}
# Define parameter ranges
param_ranges <- list(
  N_min = 2,          # Minimum dimensions 
  N_max = 30,         # Maximum dimensions
  k0_min = 0.5,         # Minimum spring constant
  k0_max = 30,        # Maximum spring constant 
  cooling_rate_min = 1e-4, # Minimum decay rate
  cooling_rate_max = 0.05, # Maximum decay rate
  c_repulsion_min = 1e-4,    # Minimum repulsion constant
  c_repulsion_max = 0.05     # Maximum repulsion constant
)

# Setup optimization parameters
opt_params <- list(
  mapping_max_iter = 1000,    # Maximum map optimization iterations
  folds = 20,         # Number of CV folds
  num_samples = 50,
  max_cores = NULL,           # Use NULL to let the function decide the number of cores
  write_files = FALSE   # Set TRUE to save intermediate results
)
```

## Initial Parameter Search

We'll run some optimization trials for each dimension and matrix variant to build an initial estimate of likelihood in the parameter space:

```{r eval=FALSE}
# Run optimization for each dimensionality and matrix variant
for(i in seq_along(results$matrices)) { 
  dim_data <- results$matrices[[i]]
  ndim <- results$metadata$dimension[i]
  
  for(matrix_type in matrix_variants) { 
    # Get current matrix
    distance_matrix <- dim_data[[matrix_type]]
    
    # Create scenario name incorporating dimension and matrix type
    scenario_name <- sprintf("synthetic_AMC203_dim%d_%s", ndim, matrix_type)
    
    message(sprintf("\nOptimizing for %d dimensions, matrix type: %s", 
                   ndim, matrix_type))
    
    # Run optimization
    opt_result <- initial_parameter_optimization(
      distance_matrix = distance_matrix,
      mapping_max_iter = opt_params$mapping_max_iter,
      relative_epsilon = 1e-4,
      convergence_counter = 5,
      scenario_name = scenario_name,
      N_min = param_ranges$N_min, 
      N_max = param_ranges$N_max,
      k0_min = param_ranges$k0_min,
      k0_max = param_ranges$k0_max, 
      c_repulsion_min = param_ranges$c_repulsion_min,
      c_repulsion_max = param_ranges$c_repulsion_max,
      cooling_rate_min = param_ranges$cooling_rate_min,
      cooling_rate_max = param_ranges$cooling_rate_max,
      folds = opt_params$folds,
      num_samples = opt_params$num_samples,
      write_files = opt_params$write_files,
      max_cores = opt_params$max_cores,
      verbose = TRUE
    )
  }
}
```

## Aggregate Results 

After the optimization jobs complete, aggregate the results for each dimension and matrix type:

```{r eval=FALSE}
# Initialize results storage
all_results <- list()

# Aggregate results for each dimension and matrix type
for(ndim in results$metadata$dimension) { 
  dim_results <- list()
  
  for(matrix_type in matrix_variants) { #
    scenario_name <- sprintf("synthetic_AMC203_dim%d_%s", ndim, matrix_type)
    print(scenario_name)
    # Aggregate optimization results
    dim_results[[matrix_type]] <- aggregate_parameter_optimization_results(
      scenario_name = scenario_name,
      write_files = TRUE
    )
  }
  
  all_results[[as.character(ndim)]] <- dim_results
}
```

## Adaptive Monte Carlo Sampling

Perform adaptive sampling for each case.
Beware that each call to run_adaptive_sampling() deletes all the files in adaptive sampling directory and uses this directory for storing its temporary files. Therefore, do not run multiple samplings in one directory concurrently.
You can check the console outputs, warnings, or errors in .out and .err files in /adaptive_sampling_jobs directory.

```{r eval=FALSE}
# Setup adaptive sampling parameters
amc_params <- list(
  num_parallel_jobs = 200,  # For local execution: Number of CPU cores available; 
  num_samples = 200,        # Number of new samples to add to the parameter distribution through AMC (for refinement.)
  mapping_max_iter = 1000,    # Maximum iterations per map optimization
  relative_epsilon = 1e-4
)

# Setup optimization parameters
opt_params <- list(
  folds = 20         # Number of CV folds

)

# Run adaptive sampling for each dimension and matrix type
for(ndim in results$metadata$dimension) { # 
  dim_data <- results$matrices[[which(results$metadata$dimension == ndim)]]
  
  for(matrix_type in matrix_variants) { # 
    # Get scenario names
    base_scenario <- sprintf("synthetic_AMC202_dim%d_%s", ndim, matrix_type)
    amc_scenario <- base_scenario #sprintf("%s_amc", base_scenario)
    
    # Get result file path
    result_file <- file.path("model_parameters", 
                            sprintf("%s_model_parameters.csv", base_scenario))
    log_transform_parameters(result_file)
    
    # Get corresponding distance matrix
    distance_matrix <- dim_data[[matrix_type]]
    
    message(sprintf("\nRunning adaptive sampling for %d dimensions, matrix type: %s", 
                   ndim, matrix_type))
    
    # Run adaptive sampling for current scenario (K-fold cross-validation happens inside) 
    run_adaptive_sampling(
      initial_samples_file = result_file,
      distance_matrix = distance_matrix,
      mapping_max_iter = amc_params$mapping_max_iter,
      scenario_name = amc_scenario,
      num_parallel_jobs = amc_params$num_parallel_jobs,
      num_samples = amc_params$num_samples,
      max_cores = opt_params$folds+1, # Each job on SLURM can work with 20 cores for a 20-fold CV
      relative_epsilon = amc_params$relative_epsilon,
      folds = opt_params$folds,
      verbose = FALSE
    )
  }
}
```

## Analyze fitted parameters Across Dimensions and Matrix Types

After all jobs complete, analyze parameter variations:

```{r}
# Initialize results collection
final_params <- data.frame()

# Collect results for each dimension and matrix type
for(ndim in results$metadata$dimension) {
  for(matrix_type in matrix_variants) {
    # Define pattern to match multiple AMC result files
    file_pattern <- sprintf("synthetic_AMC\\d+_dim%d_%s_model_parameters\\.csv", 
                           ndim, matrix_type)
    
    # Find all matching files in the model_parameters directory
    amc_files <- list.files(path = "model_parameters", 
                          pattern = file_pattern,
                          full.names = TRUE)
    
    # Check if any files were found
    if(length(amc_files) == 0) {
      cat("Warning: No files found matching pattern", file_pattern, "in model_parameters directory\n")
      next
    }
    
    cat("Found", length(amc_files), "files for dimension", ndim, "and matrix type", matrix_type, ":\n")
    cat(paste(" -", basename(amc_files)), sep = "\n")
    
    # Try to read and combine all matching files
    tryCatch({
      # Combine chain results
      amc_results <- do.call(rbind, lapply(amc_files, function(file) {
        tryCatch({
          df <- read.csv(file)
          # Add source file information for debugging
          df$source_file <- basename(file)
          return(df)
        }, error = function(e) {
          cat("Error reading file", file, ":", e$message, "\n")
          return(NULL)
        })
      }))
      
      # Remove NULL results from failed file reads
      amc_results <- amc_results[!sapply(amc_results, is.null)]
      
      if(nrow(amc_results) == 0) {
        cat("Warning: No valid data found in files for dimension", ndim, "and matrix type", matrix_type, "\n")
        next
      }
      
      # Ensure numeric columns are properly converted
      numeric_columns <- c("log_N", "log_k0", "log_cooling_rate", "log_c_repulsion", "NLL", "Holdout_MAE")
      for(col in numeric_columns) {
        if(col %in% names(amc_results)) {
          amc_results[[col]] <- as.numeric(as.character(amc_results[[col]]))
        }
      }
      
      # Filter valid results
      amc_results <- amc_results %>%
        filter(is.finite(NLL) & is.finite(Holdout_MAE)) %>%
        na.omit()
      
      # Additional filtering for valid dimensions
      amc_results <- amc_results[amc_results$log_N >= log(2),]
      
      # Clean numeric data
      numeric_cols <- sapply(amc_results, is.numeric)
      amc_results[numeric_cols] <- lapply(amc_results[numeric_cols], clean_data, k = 3.5)
      amc_results <- na.omit(amc_results)
      
      # Find best parameters (minimum NLL)
      if(nrow(amc_results) > 0) {
        # Get best parameters by minimum MAE
        best_params <- amc_results[which.min(amc_results$Holdout_MAE),]
        best_params$true_dim <- ndim
        best_params$matrix_type <- matrix_type
        best_params$N <- round(exp(best_params$log_N))
        
        # Also record file source for traceability
        best_params$source_file <- best_params$source_file
        
        final_params <- rbind(final_params, best_params)
        
        cat("Found best parameters for dimension", ndim, "matrix type", matrix_type, 
            "from file", best_params$source_file, ":\n")
        cat("  N =", best_params$N, 
            "k0 =", round(exp(best_params$log_k0), 4),
            "c_repulsion =", formatC(exp(best_params$log_c_repulsion), format = "e", digits = 2),
            "cooling_rate =", formatC(exp(best_params$log_cooling_rate), format = "e", digits = 2),
            "MAE =", round(best_params$Holdout_MAE, 4), "\n\n")
      } else {
        cat("Warning: No valid results after cleaning for dimension", ndim, 
            "and matrix type", matrix_type, "\n\n")
      }
    }, error = function(e) {
      cat("Error processing files for dimension", ndim, "and matrix type", matrix_type, ":", 
          e$message, "\n\n")
    })
  }
}

# Display summary of collected parameters
if(nrow(final_params) > 0) {
  #print(final_params[, c("true_dim", "matrix_type", "N", "Holdout_MAE")])
  
  # Save consolidated results
  results_file <- "optimal_parameters_250points_AMC10X-20X.csv"
  write.csv(final_params, results_file, row.names = FALSE)
  cat("\nSaved consolidated results to", results_file, "\n")
} else {
  cat("No valid parameters found across all dimensions and matrix types.\n")
}


cat("\nSummary of best parameters found:\n")
# Apply the transformation
final_params <- final_params %>%
  mutate(
    Missing = case_when(
      grepl("^S_", matrix_type) ~ "70%",
      grepl("^M_", matrix_type) ~ "85%",
      grepl("^L_", matrix_type) ~ "95%"
    ),
    Variant = case_when(
      grepl("_missing$", matrix_type) ~ "Original",
      grepl("_noise1$", matrix_type) ~ "+Noise",
      grepl("_noise_bias$", matrix_type) ~ "+Noise+Bias"
    )
  ) %>%
  rename("True Dim" = true_dim)

# Create a table with the requested columns and format integers without decimals
library(xtable)
latex_table <- xtable(final_params[, c("Missing", "Variant", "True Dim", "N")],
                      digits = c(0, 0, 0, 0, 0))  # 0 decimal places for all columns
print(latex_table, include.rownames = FALSE)
```

## Convergence Analysis

Check convergence for each case:

```{r}
# A helper function:
analyze_convergence <- function(ndim, matrix_type, mutual_size) {
  # Create the two specific file patterns we want to match
  file_patterns <- c(
    sprintf("synthetic_AMC101_dim%d_%s_model_parameters.csv", ndim, matrix_type),
    sprintf("synthetic_AMC102_dim%d_%s_model_parameters.csv", ndim, matrix_type),
    sprintf("synthetic_AMC201_dim%d_%s_model_parameters.csv", ndim, matrix_type),
    sprintf("synthetic_AMC202_dim%d_%s_model_parameters.csv", ndim, matrix_type),
    sprintf("synthetic_AMC203_dim%d_%s_model_parameters.csv", ndim, matrix_type)
  )

  # Get full paths for both files
  chain_files <- file.path("model_parameters", file_patterns)

  # Verify files exist
  if (!all(file.exists(chain_files))) {
    missing_files <- chain_files[!file.exists(chain_files)]
    warning("Missing chain files: ", paste(missing_files, collapse = ", "))
    chain_files <- chain_files[file.exists(chain_files)]
    if (length(chain_files) == 0) {
      stop("No chain files found for dimension ", ndim, " and matrix type ", matrix_type)
    }
  }

  # Calculate diagnostics
  diag <- calculate_diagnostics(chain_files, mutual_size = mutual_size)

  # Create diagnostic plots
  plot_file <- sprintf("synthetic_dim%d_%s_diagnostics.png", ndim, matrix_type)
  p <- create_diagnostic_plots(chain_files, mutual_size = mutual_size,
                             output_file = plot_file)

  return(list(diagnostics = diag, plot = p))
}

# Set the mutual_size, smaller then the minimum simulaations for each scenario:
mutual_size <- 400

# Analyze convergence for each case
convergence_results <- list()
for(ndim in results$metadata$dimension) {
  dim_results <- list()
  for(matrix_type in matrix_variants) {
    dim_results[[matrix_type]] <- analyze_convergence(ndim, matrix_type, mutual_size=mutual_size)
  }
}

# Print summary of convergence results
for(ndim in names(convergence_results)) {
  cat(sprintf("\nConvergence Summary for %s dimensions:\n", ndim))
  for(matrix_type in names(convergence_results[[ndim]])) {
    cat(sprintf("\nMatrix type: %s\n", matrix_type))
    print(convergence_results[[ndim]][[matrix_type]]$diagnostics)
  }
}
```

## Session Info

```{r}
sessionInfo()
```
